A tRNA modification in Mycobacterium tuberculosis facilitates optimal intracellular growth

Version 1

Abstract

Abstract

Diverse chemical modifications fine-tune the function and metabolism of tRNA. Although tRNA modification is universal in all kingdoms of life, profiles of modifications, their functions, and physiological roles have not been elucidated in most organisms including the human pathogen, Mycobacterium tuberculosis (Mtb), the causative agent of tuberculosis. To identify physiologically important modifications, we surveyed the tRNA of Mtb, using tRNA sequencing (tRNA-seq) and genome-mining. Homology searches identified 18 candidate tRNA modifying enzymes that are predicted to create 13 tRNA modifications across all tRNA species. Reverse transcription-derived error signatures in tRNA-seq predicted the sites and presence of 9 modifications. Several chemical treatments prior to tRNA-seq expanded the number of predictable modifications. Deletion of Mtb genes encoding two modifying enzymes, TruB and MnmA, eliminated their respective tRNA modifications, validating the presence of modified sites in tRNA species. Furthermore, the absence of mnmA attenuated Mtb growth in macrophages, suggesting that MnmA-dependent tRNA uridine sulfation contributes to Mtb intracellular growth. Our results lay the foundation for unveiling the roles of tRNA modifications in Mtb pathogenesis and developing new therapeutics against tuberculosis.

Introduction

tRNA is an adaptor molecule that enables protein synthesis by converting the triplet genetic code in mRNA into amino acids. The fidelity of base paring between mRNA codons and tRNA anticodons is monitored within ribosomes and is critical for properly incorporating the amino acids bound to the 3’ ends of tRNAs into growing polypeptides. For optimal translation, the abundances, and properties of tRNA isoacceptors are fine-tuned by diverse mechanisms, including chemical modifications[1-4] Dysregulation of tRNA abundance and/or structure leads to defective decoding and results in ribosome pausing and collisions, protein misfolding, stress responses and can have detrimental or lethal effects on the cell[5-10].

Chemical modifications of tRNA (tRNA modifications) are found in all kingdoms of life and fine-tune tRNA properties including mRNA decoding efficiency, recognition by aminoacyl-tRNA synthetases, half-life and structural stability[2, 10-12]. Modifications are prevalent in the anticodon loop, particularly at the first letter of the anticodon. Modifications of the anticodon loop directly modulate codon recognition, whereas modifications in the tRNA body region primarily stabilize tRNA tertiary structure, protecting them from degradation in the cell. tRNA modifications are generated by dedicated site-specific enzymes referred to as tRNA modifying enzymes. tRNA modifications have been extensively characterized in a few model organisms[2, 13], but their profiles, regulation, and functions in non-model organisms, including bacterial pathogens, are understudied[13]

tRNA sequencing (tRNA-seq) allows for the rapid and systematic prediction of many tRNA modification sites[14, 15]. We recently developed a comparative tRNA-seq protocol to profile tRNA modifications in organisms with uncharted tRNA modification profiles; in Vibrio cholerae, this approach led to the discovery of a new RNA modification and RNA editing process[16]. tRNA-seq enables rapid prediction of modified sites through detection of reverse transcription-derived signatures, such as nucleotide misincorporation and early termination, both of which occur more frequently at modified sites. Furthermore, several chemical treatments of tRNA can convert modifications that are not recognizable as reverse transcription-derived signatures into detectable signals, expanding the repertoire of modifications that can be distinguished by tRNA-seq[17-20].

Mycobacterium tuberculosis (Mtb), the agent of tuberculosis (TB), is a global pathogen that caused >10.5 million cases and over 1.5 million deaths worldwide in 2020[21]. Several studies have uncovered roles for non-Mtb mycobacterial tRNA modifications in stress responses, adaptation to environmental changes, and persister formation[22]. Mycobacterium bovis BCG, an organism closely related to Mtb, responds to hypoxia by reprogramming 40 ribonucleoside modifications in tRNA to facilitate translation of a subset of proteins that promote survival in hypoxic conditions[22]. To date, studies of the profiles and functions of Mtb tRNA modifications have been limited.

Here, we conducted tRNA-seq in Mtb. We assigned modifications to many of the reverse transcription-derived signatures identified, using information on the presence of homologs of known modifying enzymes in Mtb. Chemical treatments of tRNAs carried out prior to tRNA-seq increased the detectability of certain modifications. We constructed two Mtb deletion mutant strains, with deletions of mnmA and truB, and confirmed that the absence of these modification enzymes eliminated the predicted signals in tRNA-seq data. Furthermore, while deletion of mnmA in Mtb did not affect the pathogen’s growth in in vitro laboratory growth conditions, the mnmA knockout strain’s growth was attenuated in a macrophage infection model. Our findings suggest that tRNA modifications warrant further study as we unravel the complexity of Mtb infections, as they may serve as targets for new therapeutics.

Results

In silico prediction of Mtb tRNA modifying enzymes

To predict tRNA modifications in Mtb, we used Basic Local Alignment Search Tool (BLAST)[23] to identify homologues of all tRNA modification enzymes registered in Modomics in the Mtb genome[24]. With a stringent threshold (E-value <1×10−10), 20 Mtb genes homologous to genes encoding known RNA modification enzymes were identified. 18 of these genes are predicted to synthesize 13 tRNA modifications in Mtb (Supplementary Table 1), including miaA and miaB for 2-methylthio-6-isopentenyl-adenosine (ms2i6A), tsaD, tsaB, tsaE, and sua5 for N6 threonylcarbamoyladenosine (t6A), mnmA for 2-thiouridine (s2U), truB, truA, and pus9 for pseudouridine (Ψ), trmD for 1-methylguanosine (m 1G), trmI for 1-methyladenosine (m1A), trmL for 2’-O-methylcytidine (Cm) or 2’-O-methyluridine (Um), trmB for 7-methylguanosine (m7G), trmH for 2’-O-methylguanosine (Gm), dusB for dihydrouridine (D), tadA for inosine (I), and tilS for lysidine (k2C). While two additional genes, Rv1713 and Rv2338c, met the threshold for homology to modification enzymes, they exhibited greater similarity to a ribosome associated GTPase (Der) and a molybdopterin biosynthesis protein (MoeW) respectively, and likely do not correspond to tRNA modification enzymes.

We mined data from genome-wide Tn-seq and CRISPRi screens[25, 26] to assess the impacts of Mtb tRNA modifications on its growth. Five genes encoding Mtb tRNA modifying enzymes, trmD, tilS, tadA, sua5, and miaA, were reported to be essential for Mtb growth in both Tn-seq[25] and CRISPRi screens[26] (Supplementary Table 1), suggesting that the modifications they produce are critical for tRNA functions. Indeed, E. coli trmD, tilS, tadA, and sua5 are also essential for growth and critical for codon decoding, aminoacylation, and reading frame maintenance[27-30]. The Mtb modifications synthesized by these enzymes likely have similar impacts on tRNA functions. Unexpectedly, one modifying enzyme, MiaA, is non-essential in E. coli, but apparently essential in Mtb, suggesting that i6A, the modification introduced by MiaA, may have more profound roles in Mtb translation than in E. coli.

Several of the putative Mtb tRNA modifying enzymes are conserved across all three domains of life (e.g., TruB and TsaD) (Fig. 1). By contrast, some enzymes were limited to bacterial species closely related to Mtb, possibly suggesting their species-specific physiological roles, including pathogenesis. For example, TrmI (Rv2118c) homologs are widely present in Archaea and Eukaryotes, but are sparsely distributed in bacteria, where they are primarily limited to Actinomycetia, including Mtb, and several thermophilic bacterial species (e.g., Thermus thermophilus). TrmI synthesizes m1A at position 58 in eukaryotes[31] and at position 57/58 in archaea[32]; indeed, TrmI in Mtb has been proven to generate m1A at position 58[33]. Pus9 (Rv3300c) has many homologs exhibiting weak similarity across eukaryotes and bacteria. However, some species, including Pseudomonas aeruginosa, Acinetobacter baumannii, Neisseria gonorrhoeae, and several species of Actinomycetia, encode proteins showing strong homology to Mtb Pus9, suggesting that this set of pseudouridylases have a distinctive property such as substrate specificity.

Phylogenetic distribution of Mtb tRNA modifying enzyme homologs.

Heat map of log10 E-values from BLAST search results. BLAST searches were conducted against 118 manually picked organisms using Mtb tRNA modifying enzymes as queries. When one organism has multiple hits, the lowest log10(Eval) values among hits is shown. iTol[34] was used to depict the results.

Profiling Mtb tRNA modification sites by tRNA sequencing

To begin profiling detectable Mtb tRNA modifications, we sequenced tRNAs isolated from wild type Mtb strain H37Rv grown in 7H9 medium. In this protocol, tRNAs are first reversed transcribed to cDNA[16]. During cDNA synthesis, chemical modifications on tRNA nucleotides disrupt Watson-Crick base pairing and increase the frequency of reverse transcriptase errors, leading to incorporation of the incorrect nucleotide or early termination of cDNA synthesis[35]. These reverse transcription derived ‘signatures’ typically correspond to modified sites[14, 15] and are depicted in the heat map in Fig. 2.

Heat map of misincorporation and early termination frequency in sequencing of tRNAs from wild type Mtb.

(A, B) Heatmaps show misincorporation (A) and termination (B) frequencies at all positions across tRNAs (read 5’ to 3’). Predicted modifications are labeled based on similarity to known modifications in other organisms and the presence of the tRNA modifying enzyme homologs (Supplementary Table 1). The positions with more than 10% misincorporation in Mtb but not in E. coli are depicted in white in A. (C) M. tuberculosis tRNA modifications predicted in this study. Schematic tRNA secondary structure with sites of modifications identified either by the presence of modifying enzymes and/or tRNA-seq. Modifications and tRNA species that are not observed in E. coli are shown in red. Modifications and positions that are predicted by both RT-derived signature and the presence of the homologs of tRNA modifying enzymes are shown in yellow (without chemical treatment) and green (with chemical treatment), whereas modifications that are only predicted by the presence of the homologs are shown in light blue. Genes reported to be essential in Mtb are shown in bold.

Comparison of the reverse transcription derived signatures observed in Mtb to E. coli, in which tRNA modifications are previously well characterized[16] enables the prediction of the presence of common modifications, including ms2i6A, m1G, I, and k2C (Fig 2). These predictions are strongly supported by the set of tRNA modification enzymes identified in the Mtb genome (Fig. 1 and Supplementary Table 1), including miaA and miaB (ms2i6A), trmD (m1G), tadA (I), and tilS (k2C). Some tRNA modifications were observed in Mtb but are not present in E. coli. In other actinobacteria, A58 and A59 are likely modified to m1A[36], and since trmI, the methylase that generates this modification is present in Mtb[33], most Mtb tRNAs likely contain this modification as well. Nucleoside variation in the sequence of tRNA genes can account for some of the variations in modified sites between E. coli and Mtb. For example, in Mtb, termination signatures derived from G at position 37 were detected in tRNA-Arg2, -Gln1, and -Gln2, whereas this position in these tRNAs in E. coli are modified A, such as m2A and m6A, which are silent in tRNA-seq. Since m1G induces strong termination during reverse transcription, these positions are likely modified to m1G, as observed in the Bacillus subtilis tRNA-Arg2 gene position 37 G[37].

tRNA samples were also treated with several chemical treatments prior to sequencing, to expand the set of tRNA modifications detectable by tRNAseq. These treatments included iodoacetamide (IAA), for detection of sulfur modifications,1-cyclohexyl-(2-morpholinoethyl) carbodiimide (CMC) for detection of Ψ, and alkali for detection of D and m7G. The chemical treatment protocols were first carried out with E. coli, to validate the methods.

IAA is a thiol-reactive compound that covalently attaches carboxyamidomethyl to thiolated uridines via nucleophilic substitution[38] and modified s4U is detected as C instead of U. In IAA-treated samples, position 8 and 9, corresponding to s4U in many tRNA species, had high misincorporation frequencies, confirming that IAA treatment modifies s4U, leading to elevated misincorporation (Supplementary Fig. 1). Furthermore, we observed higher misincorporation and termination signals at the positions corresponding to other sulfur modification, s2C, s2U and their derivatives, such as position 32 in tRNA-Arg3, -Arg5, -Ser3, and Arg4, and 34 in tRNA-Glu, -Gln1, and -Lys (Fig. 3 and Supplementary Fig. 1), revealing that IAA treatment facilitates the detection of not only s4U but also additional sulfur modifications, which are weakly detected without the IAA treatment.

IAA treatment promotes detection of sulfur modifications on tRNAs by enhancing termination signals.

(A, B) Heatmaps of the termination signals of E. coli tRNAs treated with (A) or without (B) IAA. Known modification sites, including sulfur modifications (s4U, s2C, s2U in white) are shown. (C) Termination frequency at s 4U, s2C, and s2U sites of tRNAs treated with or without IAA.

Next, we applied IAA treatment to Mtb tRNA-seq. IAA treatment increased termination signals from position 34 in tRNA-Glu1, -Gln1, and -Lys, which contain s2U derivatives in E. coli (Fig. 4A). The 2-thiouridine modification is carried out by MnmA in E. coli[39], and a homolog, Rv3024c, of this enzyme was identified in the Mtb genome (Supplementary Table 1)[40]. We used double stranded DNA-based recombineering to delete Rv3024c in Mtb, yielding strain MtbΔmnmA[41]. Sequencing of tRNA isolated from MtbΔmnmA with prior IAA treatment showed reduced termination signals from position 34 in tRNA-Glu1, -Gln1, and -Lys in treated samples, indicating that Rv3024c plays a critical role in the modification responsible for increased termination frequency derived from position 34 in these tRNAs (Fig. 4). Together, these observations strongly suggest that Rv3024c encodes an MnmA-like enzyme that sulfurates position 34 uridines in three Mtb tRNA isoacceptors.

Heat map and plot of early termination frequency from sequencing of tRNAs from wild type and MtbΔmnmA with and without RNA alkylation.

(A, B) Heat map of early termination frequencies across tRNA molecules and positions for WT (A) and MtbΔmnmA (B). Sulfur modification are shown in white. (B) Plot of termination frequencies at position 37 in wild type (WT) Mtb and MtbΔmnmA for lysine_UUU, glutamate_UUG, and glutamine_UUC isoacceptors. IAA; iodoacetamide.

As shown previously[42], CMC treatment increased both misincorporation and termination signatures at a subset of Ψs in E. coli tRNAs (Fig 5, and Supplementary Fig. 2 and 3). In addition, CMC-treated samples showed increased frequencies of both misincorporation and termination at position 16, 17, 20, and 20A corresponding to D, and m7G at position 46. Both modifications are known to undergo base elimination in mild alkali conditions[43]. Since these signals were also observed in the reaction condition in which CMC was not added, these signals are likely attributable to the alkali treatment that is common to both conditions.

CMC and alkali treatment facilitate detection of Ψ, D, and m7G modifications in E. coli.

(A, B) Heatmaps of the misincorporation signals of E. coli tRNAs treated with (A) or without (B) CMC. In both conditions, tRNAs are incubated in alkali condition. Known Ψ, D, and m7G sites are shown. Ψ are shown in white. (C) Misincorporation frequency at known Ψ, D, m7G sites of tRNAs treated with CMC+/alkali+ or CMC-/alkali+, and tRNAs without treatment.

Reverse transcription-derived signatures derived from alkali-treated D showed a distinctive pattern. With this treatment, when two Ds are at consecutive positions, e.g., D16 and D17, termination signals were elevated at the following position (Supplementary Fig. 2 and 4). Furthermore, alkali treatment also led to higher misincorporation frequencies at singlet Ds (Fig. 5 and Supplementary Fig. 4). Thus, termination and misincorporation signatures enabled the prediction of known E. coli tRNA sites modified to D.

CMC/alkali treatment facilitated the identification of additional modifications in Mtb tRNAs. Regardless of CMC treatment, alkali treated samples showed increased misincorporation and termination frequencies derived from U located at position 16, 17, 20, and 20A (Fig. 6 and Supplementary Fig. 5). As observed in E. coli, termination signals at position 18 and 21 likely correspond to consecutive Ds at position 16 and 17, and 20 and 20A, respectively. Mtb Rv0823c is a homolog of dihydrouridylase DusB (Supplementary Table 1), which likely accounts for the synthesis of D at these positions. Furthermore, alkali treatment increased the misincorporation frequencies at G at position 46 (Supplementary Fig. 6). Since Rv0208c is a homolog of TrmB, which synthesizes m7G at position 46 in E. coli, multiple Mtb tRNA species likely contain m7G at position 46 (Fig. 6 and Supplementary Fig. 6).

Heat map of early termination frequency from sequencing of tRNAs isolated from WT and MtbΔtruB following CMC treatment.

Heat map of early termination frequencies across tRNA molecules and positions for WT (left) and MtbΔtruB (right). Termination signals derived from position 55 are shown in white.

CMC treatment also increased the termination frequency at several sites. Termination signatures derived from position 55 increased in most tRNA species, suggesting that Mtb tRNAs contain pseudouridines at this position[37]. Rv2793c is an Mtb homologue of E. coli TruB and deletion of Rv2793c reduced the termination frequencies at this position in tRNAs isolated from MtbΔtruB (Fig. 6 and Supplementary Fig. 7)[41]. Together, these observations suggest that Rv2793c encodes a TruB-like enzyme that modifies position 55 uridines to Ψ across tRNA species. Furthermore, the presence of a TruA homolog in Mtb (Rv3455c) suggests that in multiple tRNA species U at positions 38-40 can be modified to Ψ. Indeed, the termination signatures derived from positions 38 and 39 increased depending on CMC treatment, strongly suggesting that these positions are modified to Ψ (Fig. 6 and Supplementary Fig. 7).

In total, among 13 tRNA Mtb tRNA modifications predicted by the presence of tRNA modifying enzymes (Fig 1 and Supplementary Table 1), 9 species of modifications were detected based on reverse-transcription derived signatures (Fig. 2C).

Growth of MtbΔmnmA is attenuated in a macrophage infection model

To address whether Mtb tRNA modifications impact the pathogen’s growth in the host environment, we used the MtbTnDB transposon insertion sequencing (Tn-seq) database[44, 45] to determine if transposon insertions in genes encoding tRNA modifying enzymes have been associated with in vivo growth defects. Transposon insertions in mnmA were reported to attenuate Mtb growth in mice infected with a library of Mtb transposon mutants, suggesting that mnmA facilitates Mtb growth in vivo. We found that the growth of WT and MtbΔmnmA were similar in 7H9 medium (Fig. 7), suggesting the absence of s2U modification at position 34 does not impair Mtb growth in culture. In contrast, the MtbΔmnmA mutant was significantly impaired for growth in a macrophage infection model[45] (Fig. 7). Defective growth of the MtbΔmnmA mutant was also observed in macrophages treated with all-trans retinoic acid (ATRA), which promotes macrophage control of Mtb infection[46]. These observations strongly suggest that modification of U to s2U by MnmA facilitates Mtb growth in macrophages.

MtbΔmnmA is attenuated in a macrophage infection model.

Top: Wild type and MtbΔmnmA do not display growth differences in 7H9 medium. Bottom: Auto-luminescent wild type and ΔmnmA Mtb strains were diluted to a multiplicity of infection of 2 bacteria per mouse bone marrow-derived macrophage with or without all-trans retinoic acid (ATRA). Survival was measured by luminescence and normalized to luminescence at time 0.

Discussion

Here, we profiled Mtb tRNA modifications with tRNA sequencing to provide the first maps of the tRNA modification landscape in this global pathogen. In total, nine modifications, including six modifications without chemical treatment, were identified based on reverse transcription derived signatures. CMC/alkali treatment and IAA treatment further identified Ψ and m7G and sulfur modifications, respectively. Although we did not chemically validate the modifications predicted by tRNA-seq with mass spectrometry, the identification of Mtb homologs of tRNA modifying enzymes strongly bolsters the RT-signature-based predictions. Furthermore, the deletion of truB and mnmA genes in Mtb eliminated the respective modification signatures of pseudouridine and s2U, validating that the enzymes encoded by these genes synthesize these modifications. Finally, the growth defect of the ΔmnmA strain within macrophages but not in a nutrient-rich medium suggests that s2U tRNA modification facilitates Mtb adaptation to the host intracellular environment.

IAA treatment was developed for detecting s4U modification in pulse-chase experiments to measure RNA turnover[38]. We found that this treatment enhances the RT-signatures of additional sulfur modifications as well, including s2C and s2U, indicating that IAA should have general utility in tRNA-seq-based profiling of sulfur modifications, including in studies tracking changes in tRNA sulfuration[47, 48].

Multiple modifications do not cause reverse transcriptase errors. Although tRNA-seq provides a simple and rapid method for profiling the landscape of modifications in all tRNA species, this approach does not enable comprehensive identification of modifications. Here, several modifications, including t6A, Cm, Um, and Gm, predicted by the presence of Mtb homolog of tRNA modifying enzymes, such as Sua5, TsaB, TsaD, TsaE, TrmL, and TrmH, were not detected in tRNA-seq. Mapping and further analysis of these modifications will require different approaches such as tRNA purification and RNA mass spectrometric analysis.

Several strong signatures were detected in Mtb tRNAs but not in E. coli (Fig. 2A), including G45 in tRNA-Gly2. Most of these signals are strict misincorporations of C or T at A or G position, respectively. Since modifications at a purine position, e.g., m1G and m 1A, generally cause random misincorporation of the other three nucleosides, strong Mtb-specific misincorporation signatures are not likely to be derived from modification. Further mass spectrometric analysis of purified tRNAs will be necessary to assess whether these signals are derived from Mtb-specific modifications.

The role of s2U in Mtb appears to be unusual. s2U is a universally conserved modification observed in all three domains of life at the wobble position in the anticodon. This modification enhances the stacking of s2U with U35 to stabilize the anticodon structure, facilitating codon-anticodon interactions. s2U is also recognized by multiple aminoacyl-tRNA synthetases for efficient amino acylation[12]. Although elimination of this sulfur modification causes severe growth phenotypes in most organisms[39, 49], unexpectedly, the deletion of mnmA in Mtb did not attenuate growth of the pathogen in vitro, suggesting that the Mtb requirement for s2U modification differs from other organisms. The marked growth retardation of ΔmnmA strain within macrophages indicates the specific requirement of this modification within host cells. This modification may be necessary for maintaining general translation efficiency inside host cells and/or facilitate the expression of specific genes that are necessary for survival within macrophages.

In most organisms, s2U is further modified into derivatives containing an additional chemical moiety at position 5[2, 50, 51]. However, Mtb does not contain apparent homologs of the tRNA modifying enzymes that introduce the additional modifications to s2U. Thus, Mtb may contain s2U or s2U derivatives synthesized by other types of enzymes. Additional analyses to elucidate the structures of modified s2U in Mtb are warranted.

Our findings serve as a valuable starting point for the research community to continue characterizing the physiological roles and mechanisms of Mtb tRNA modifications. Since tRNA-seq offers an efficient and scalable platform for surveying changes in tRNA modifications, this approach will be valuable across growth conditions, and may be extended to growth inside host cells. Finally, further studies elucidating the mechanisms by which tRNA modifications facilitate Mtb growth in host cells should be valuable for designing new therapeutics for tuberculosis.

Supporting information

eLife Assessment

This is a valuable addition to the literature as it helps us understand the role of tRNA modifying enzymes in Mycobacterium tuberculosis. By knocking out one of the enzymes, the authors convincingly demonstrate the importance of tRNA-modifying enzymes for intra-host growth of tubercle bacteria. Some of the claims regarding modification as well as the role in virulence could be strengthened through further bioinformatics and phylogenetic analyses as well as experimental approaches. The work will be of interest to microbiologists.

Reviewer #1 (Public Review):

Tomasi et al. performed a combination of bioinformatic, next-generation tRNA sequencing experiments to predict the set of tRNA modifications and their corresponding genes in the tRNAs of the pathogenic bacteria Mycobacterium tuberculosis. Long known to be important for translation accuracy and efficiency, tRNA modifications are now emerging as having regulatory roles. However, the basic knowledge of the position and nature of the modifications present in a given organism is very sparse beyond a handful of model organisms. Studies that can generate the tRNA modification maps in different organisms along the tree of life are good starting points for further studies. The focus here on a major human pathogen that is studied by a large community raises the general interest of the study. Finally, deletion of the gene mnmA responsible for the insertion of s2U at position 34 revealed defects in in growth in macrophage but in test tubes suggesting regulatory roles that will warrant further studies. The conclusions of the paper are mostly supported by the data but the partial nature of the bioinformatic analysis and absence of Mass-Spectrometry data make it incomplete. The authors do not take advantage of the Mass spec data that is published for Mycobacterium bovis (PMID: 27834374) to discuss what they find.

Important points to be considered:

1. The authors say they took a list of proteins involved in tRNA modifications from Modomics and added manually a few but we do not know the exact set of proteins that were used to search the M. mycobacterium genome.

2. The absence of mnmGE genes in TB suggested that the xcm5U derivatives are absent. These are present in M. bovis (PMID: 27834374). Are the MnmEG gene found in M. bovis? If yes, then the authors should perform a phylogenetic distribution analysis in the Mycobacterial clade to see when they disappeared. If they are not present in M. bovis then maybe a non-orthologous set of enzymes do the same reaction and then the authors really do not know what modification is present or not at U34 without LC-MS. The exact same argument can be given for the xmo5U derivatives that are also found in M.bovis but not predicted by the authors in M. tuberculosis.

3. Why is the Psi32 predicted by the authors because of the presence of the Rv3300c/Psu9 gene not detected by CMC-treated tRNA seq while the other Psi residues are? Members of this family can modify both rRNA and tRNA. So the presence of the gene does not guarantee the presence of the modification in tRNAs

4. What are tsaBED not essential but tsaC (called sua5 by the authors) essential?

Reviewer #2 (Public Review):

In this study, Tomasi et al identify a series of tRNA modifying enzymes from Mtb, show their function in the relevant tRNA modifications and by using at least one deleted strain for MnmA, they show the relevance of tRNA modification in intra-host survival and postulate their potential role in pathogenesis.

Conceptually it is a wonderful study, given that tRNA modifications are so fundamental to all life forms, showing their role in Mtb growth in the host is significant. However, the authors have not thoroughly analyzed the phenotype. The growth defect aspect or impact on pathogenesis needs to be adequately addressed.

- The authors show that ΔmnmA grows equally well in the in vitro cultures as the WT. However, they show attenuated growth in the macrophages. Is it because Glu1_TTC and Gln1-TTG tRNAs are not the preferred tRNAs for incorporation of Glu and Gln, respectively? And for some reason, they get preferred over the alternate tRNAs during infection? What dictates this selectivity?

- As such the growth defect shown in macrophages would be more convincing if the authors also show the phenotype of complementation with WT mnmA.

An important consideration here is the universal nature of these modifications across the life forms. Any strategy to utilize these enzymes as the potential therapeutic candidate would have to factor in this important aspect.

Reviewer #3 (Public Review):

The work presented in the manuscript tries to identify tRNA modifications present in Mycobacterium tuberculosis (Mtb) using reverse transcription-derived error signatures with tRNA-seq. The study identified enzyme homologs and correlates them with presence of respective tRNA modifications in Mtb. The study used several chemical treatments (IAA and alkali treatment) to further enhance the reverse transcription signals and confirms the presence of modifications in the bases. tRNA modifications by two enzymes TruB and MnmA were established by doing tRNA-seq of respective deletion mutants. Ultimately, authors show that MnmA-dependent tRNA modification is important for intracellular growth of Mtb. Overall, this report identifies multiple tRNA modifications and discuss their implication in Mtb infection.

Important points to be considered:

- The presence of tRNA-based modifications is well characterised across life forms including genus Mycobacterium (Mycobacterium tuberculosis: Varshney et al, NAR, 2004; Mycobacterium bovis: Chionh et al, Nat Commun, 2016; Mycobacterium abscessus: Thomas et al, NAR, 2020). These modifications are shown to be essential for pathogenesis of multiple organisms. A comparison of tRNA modification and their respective enzymes with host organism as well as other mycobacterium strains is required. This can be discussed in detail to understand the role of common as well as specific tRNA modifications implicated in pathogenesis.

- Authors state in line 293 "Several strong signatures were detected in Mtb tRNAs but not in E. coli". Authors can elaborate more on the unique features identified and their relevance in Mtb infection in the discussion or result section.

- Deletion of MnmA is shown to be essential for E. coli growth under oxidative stress (Zhao et al, NAR, 2021). In similar lines, MnmA deleted Mtb suffers to grow in macrophage. Is oxidative stress in macrophage responsible for slow Mtb growth?

- Authors state in line 311-312 "Mtb does not contain apparent homologs of the tRNA modifying enzymes that introduce the additional modifications to s2U". This can be characterised further to rule out the possibility of other enzyme specifically employed by Mtb to introduce additional modification.

Author Response

The following is the authors’ response to the original reviews.

We thank the editors and reviewers for their thoughtful consideration of our manuscript. Here, we addressed the reviewers’ points.

Reviewer #1 (Public Review):

Tomasi et al. performed a combination of bioinformatic, next-generation tRNA sequencing experiments to predict the set of tRNA modifications and their corresponding genes in the tRNAs of the pathogenic bacteria Mycobacterium tuberculosis. Long known to be important for translation accuracy and efficiency, tRNA modifications are now emerging as having regulatory roles. However, the basic knowledge of the position and nature of the modifications present in a given organism is very sparse beyond a handful of model organisms. Studies that can generate the tRNA modification maps in different organisms along the tree of life are good starting points for further studies. The focus here on a major human pathogen that is studied by a large community raises the general interest of the study. Finally, deletion of the gene mnmA responsible for the insertion of s2U at position 34 revealed defects in in growth in macrophage but in test tubes suggesting regulatory roles that will warrant further studies. The conclusions of the paper are mostly supported by the data but the partial nature of the bioinformatic analysis and absence of Mass-Spectrometry data make it incomplete. The authors do not take advantage of the Mass spec data that is published for Mycobacterium bovis (PMID: 27834374) to discuss what they find.

1. The authors say they took a list of proteins involved in tRNA modifications from Modomics and added manually a few but we do not know the exact set of proteins that were used to search the M. mycobacterium genome.

Thank you for pointing out this issue. We added the complete list of proteins used for the BLAST query as Supplemental Table 1.

1. The absence of mnmGE genes in TB suggested that the xcm5U derivatives are absent. These are present in M. bovis (PMID: 27834374). Are the MnmEG gene found in M. bovis? If yes, then the authors should perform a phylogenetic distribution analysis in the Mycobacterial clade to see when they disappeared. If they are not present in M. bovis then maybe a non-orthologous set of enzymes do the same reaction and then the authors really do not know what modification is present or not at U34 without LC-MS. The exact same argument can be given for the xmo5U derivatives that are also found in M.bovis but not predicted by the authors in M. tuberculosis.

The reviewer raises a valid point. In M. bovis mnm5U and cmo5U derivatives were observed in LC-MS analysis. However, we did not identify candidate genes known to be involved in the biogenesis of mnm5U and cmo5U in the Mycobacteriaceae, including M. bovis and Mtb, suggesting that if these modifications are indeed present, they are not synthesized through canonical biogenesis pathways in this family. There are several examples where the same modification is generated by distinct modification enzymes (Kimura, 2021). These observations raise the interesting possibility that in the Mycobacteriaceae and most species in actinomycetota (except for Bifidobacterium, Corynebacterium and Rhodococcus species), major wobble modifications are generated by biosynthesis pathways that are distinct from those employed by well-characterized organisms. Future studies will examine this hypothesis.

1. Why is the Psi32 predicted by the authors because of the presence of the Rv3300c/Psu9 gene not detected by CMC-treated tRNA seq while the other Psi residues are? Members of this family can modify both rRNA and tRNA. So the presence of the gene does not guarantee the presence of the modification in tRNAs

Thank you very much for the careful read. We did not include RluA in the list of query proteins because it is not classified as a tRNA modification enzyme in Modomics. Additionally, the CMC-coupled tRNA-seq is imperfect for detection of all pseudouridylated positions. Due to this limitation, we only assigned modifications that are both predicted by the presence of putative biosynthetic enzymes and RT-derived signatures. As the reviewer points out, we cannot rule out that this homolog targets only rRNAs. We clarified this possibility in the revised manuscript by adding the following sentence: “Additionally, CMC treatment may not identify  at all positions, thus, the targets of Rv3300 and Rv1540 remain unclear. Since these genes are similar to E. coli rluA, which also targets rRNA, these genes may target rRNAs instead of tRNAs” (lines 298-300)

In the revised manuscript, RluA was added to the BLAST query for creating Fig. 2. Interestingly, Rv3300c is more similar to Pus9 than RluA, while Rv1540 is the Mtb gene most similar to E. coli RluA suggesting that these two genes encode pseudouridylases that target different species of tRNAs/rRNAs.

1. What are tsaBED not essential but tsaC (called sua5 by the authors) essential?

Thank you for pointing out this interesting observation. We are also curious about differences in the essentiality among t6A biogenesis genes. We speculate that TsaC has critical roles in cell viability other than t6A synthesis. TsaC synthesizes threonylcarbamoyl-AMP as an intermediate for t6A biogenesis. Thus, it is possible that this intermediate has a role in other essential cellular activities besides t6A biogenesis. Further study of these factors in Mtb could reveal interesting crosstalk between modification synthesis and other cellular activities.

Reviewer #2 (Public Review):

In this study, Tomasi et al identify a series of tRNA modifying enzymes from Mtb, show their function in the relevant tRNA modifications and by using at least one deleted strain for MnmA, they show the relevance of tRNA modification in intra-host survival and postulate their potential role in pathogenesis.

Conceptually it is a wonderful study, given that tRNA modifications are so fundamental to all life forms, showing their role in Mtb growth in the host is significant. However, the authors have not thoroughly analyzed the phenotype. The growth defect aspect or impact on pathogenesis needs to be adequately addressed.

The authors show that ΔmnmA grows equally well in the in vitro cultures as the WT. However, they show attenuated growth in the macrophages. Is it because Glu1_TTC and Gln1-TTG tRNAs are not the preferred tRNAs for incorporation of Glu and Gln, respectively? And for some reason, they get preferred over the alternate tRNAs during infection? What dictates this selectivity?

Thank you very much for raising this excellent point. As the reviewer suggests, the attenuation of ΔmnmA Mtb growth inside of macrophages could be caused by disparate codon usage between genes required for in vitro growth and intracellular growth. Among multiple codons encoding Glu, Gln, or Lys, s2U modification-dependent codons might be preferentially distributed in genes associated with intracellular growth. For example, Mtb has two tRNA isoacceptors, Glu1_TTC and Glu2_CTC, to decipher two Glu codons, GAA and GAG. According to the wobble pairing rule, GAA is only decoded by Glu1_TTC, whereas GAG is decoded by both Glu1_TTC and Glu2_CTC; i.e., GAG can be deciphered by an s2U-independent tRNA. Thus, genes required for intracellular growth might be enriched with GAA, an s2U-dependent codon. Similar codon usage differences could be present in Gln and Lys codons deciphered by s2U-containing tRNAs. In the revised manuscript, we included a new paragraph in the discussion explaining the possibility that differences in codon usage could contribute to the intracellular fitness defect of the ΔmnmA Mtb mutant (lines 323-332).

As such the growth defect shown in macrophages would be more convincing if the authors also show the phenotype of complementation with WT mnmA.

The reviewer raises a valid point. We note however, that Rv3023c, a putative transposase, is downstream of MnmA and unlike MnmA, Rv3023c appears to be dispensable for in vivo growth, according to the Tn-seq database (reference 44 and 45). Therefore, it is likely that the intracellular growth defect is caused by loss of mnmA.

An important consideration here is the universal nature of these modifications across the life forms. Any strategy to utilize these enzymes as the potential therapeutic candidate would have to factor in this important aspect.

This is a valid point. Targeting a pathogen-specific system enables avoidance of the adverse side effects caused by many therapeutic reagents. There are a couple of Mtb modification enzymes that are specific to bacteria and critical for Mtb fitness (e.g., TilS). These enzymes represent ideal potential therapeutic targets to impede Mtb intracellular growth.

Reviewer #3 (Public Review):

The work presented in the manuscript tries to identify tRNA modifications present in Mycobacterium tuberculosis (Mtb) using reverse transcription-derived error signatures with tRNA-seq. The study identified enzyme homologs and correlates them with presence of respective tRNA modifications in Mtb. The study used several chemical treatments (IAA and alkali treatment) to further enhance the reverse transcription signals and confirms the presence of modifications in the bases. tRNA modifications by two enzymes TruB and MnmA were established by doing tRNA-seq of respective deletion mutants. Ultimately, authors show that MnmA-dependent tRNA modification is important for intracellular growth of Mtb. Overall, this report identifies multiple tRNA modifications and discuss their implication in Mtb infection.

Important points to be considered:

The presence of tRNA-based modifications is well characterised across life forms including genus Mycobacterium (Mycobacterium tuberculosis: Varshney et al, NAR, 2004; Mycobacterium bovis: Chionh et al, Nat Commun, 2016; Mycobacterium abscessus: Thomas et al, NAR, 2020). These modifications are shown to be essential for pathogenesis of multiple organisms. A comparison of tRNA modification and their respective enzymes with host organism as well as other mycobacterium strains is required. This can be discussed in detail to understand the role of common as well as specific tRNA modifications implicated in pathogenesis.

The reviewer raises a fair point. However, with the exception of Chionh et al., the other studies cited here are not genome-wide characterization of tRNA modification. Re-analysis showed that the distribution of the tRNA modifying enzymes are very similar across mycobacterium strains, e.g., Mycobacterium smegmatis, Mycobacterium tuberculosis, and Mycobacterium abscessus, suggesting that modifications related to pathogenesis in Mtb may have different physiological roles in other Mycobacterium species. We included the distribution of tRNA modification enzymes across multiple mycobacterium species in a revised Fig. 1.

Authors state in line 293 "Several strong signatures were detected in Mtb tRNAs but not in E. coli". Authors can elaborate more on the unique features identified and their relevance in Mtb infection in the discussion or result section.

Thank you for the suggestion. However, the identity of these RT signatures and the relevance of these modifications for Mtb pathogenicity remains speculative at this point.

Deletion of MnmA is shown to be essential for E. coli growth under oxidative stress (Zhao et al, NAR, 2021). In similar lines, MnmA deleted Mtb suffers to grow in macrophage. Is oxidative stress in macrophage responsible for slow Mtb growth?

This is an excellent hypothesis which we have added to the revised manuscript (lines 320-322). “In fact, the absence of mnmA is reported to sensitize E. coli to oxidative stress, raising the possibility that s2U modification promotes Mtb growth under oxidative stress elicited by the host.”

Authors state in line 311-312 "Mtb does not contain apparent homologs of the tRNA modifying enzymes that introduce the additional modifications to s2U". This can be characterised further to rule out the possibility of other enzyme specifically employed by Mtb to introduce additional modification.

The reviewer raises a valid point. As discussed above (Reviewer #1, pt 2), Mtb may employ distinct enzymes to generate certain tRNA modifications. Future mass spec-based analyses of Mtb tRNAs will be carried out to identify the precise chemical structure of the sulfurated uridine, and subsequent studies will attempt to determine the enzymes that account for the biogenesis of these modifications.

Kimura, S. (2021). Distinct evolutionary pathways for the synthesis and function of tRNA modifications. Brief Funct Genomics, 20(2), 125-134. doi:10.1093/bfgp/elaa027

Reviewer #1 (Recommendations For The Authors):

Additional data and Analyses

The Modomics database is far from complete so it would be more rigorous to give the full set of genes that was used to do the searches as supplemental data.

Thank you for the suggestion. We added the list of the query genes as Supplemental Table 1. Minor points to be fixed

1. The authors name the psi32 synthase Rv3300c Pus9 when it is a member of the RluA family. It is not clear why the yeast/eukaryotic name was used.

We included enzymes from diverse species in our query, including eukaryotic genes. Indeed, we found that Rv3300c showed the lowest E-value among our query genes, therefore, we name Rv3300c as Pus9.

1. The sua5 gene name was used it should be tsaC to follow the accepted nomenclature.

We renamed Sua5 to TsaC2.

1. The statement lines 203-296 was totally unclear. I did not understand what the authors were trying to say at all.

This paragraph described how sequence context can result in different reverse transcription-derived signatures from dihydrouridine (D). We added a schematics describing this paragraph as Supplemental Fig. 6.

1. In reference, names with special characters should be fixed such as Börk.

We fixed the names with special characters.

Reviewer #2 (Recommendations For The Authors):

The authors state that at least some of tRNA modifying enzymes, while redundant for growth in vitro, may play a role during growth inside the macrophages, mostly due to the diverse stresses they could encounter.

We added a sentence, “In fact, the absence of mnmA is reported to sensitize E. coli to oxidative stress, suggesting that s2U modification is required for Mtb growth under oxidative stress elicited by the host” in the discussion.

• Ideally, authors could have tested the impact of diverse intracellular stresses that Mtb encounters, like redox stress, nitrosative, pH or nutritional stress, to check whether any of these stresses cause in vitro growth defects in ΔmnmA strain.

Thank you for the suggestion. This point will be addressed in future experiments.

This would be a wonderful way to show that under stress, the essentiality of tRNA modification enzymes changes.

Reviewer #3 (Recommendations For The Authors):

• In general, the clarity of the presentation can be improved.

• Authors state that "MiaA, is non-essential in E. coli, but apparently essential in Mtb". While MiaA is shown to be critical for the fitness and virulence of extraintestinal pathogenic E. coli (Fleming et al, NAR, 2022). This can be clarified.

We rephrase as follows: “Unexpectedly, one modifying enzyme, MiaA, is non-essential in E. coli grown in nutrient-rich medium, but apparently …”

• Line numbers 130-132 is a repetition of line numbers 103-105

We repeated these sentences because the same claim was deduced from different experiments, i.e., BLAST search and tRNAseq.

• Line number 228: The presence of U at position 55 in the tRNAs can be included in the text for a better understanding.

We changed the text as following: “… Termination signatures derived from position 55, which is exclusively uridine in all tRNA species, increased in most tRNA species, suggesting …”

• A detailed pictorial depiction on comparing the modifications and enzymes from E. coli and Mtb can be included for easy understanding.

We created an E. coli tRNA modification map in the same format as Figure 2C and added it to the revised manuscript as a new Supplementary Fig. 1.

Version 2

Abstract

Abstract

Diverse chemical modifications fine-tune the function and metabolism of tRNA. Although tRNA modification is universal in all kingdoms of life, profiles of modifications, their functions, and physiological roles have not been elucidated in most organisms including the human pathogen, Mycobacterium tuberculosis (Mtb), the causative agent of tuberculosis. To identify physiologically important modifications, we surveyed the tRNA of Mtb, using tRNA sequencing (tRNA-seq) and genome-mining. Homology searches identified 23 candidate tRNA modifying enzymes that are predicted to create 16 tRNA modifications across all tRNA species. Reverse transcription-derived error signatures in tRNA-seq predicted the sites and presence of 9 modifications. Several chemical treatments prior to tRNA-seq expanded the number of predictable modifications. Deletion of Mtb genes encoding two modifying enzymes, TruB and MnmA, eliminated their respective tRNA modifications, validating the presence of modified sites in tRNA species. Furthermore, the absence of mnmA attenuated Mtb growth in macrophages, suggesting that MnmA-dependent tRNA uridine sulfation contributes to Mtb intracellular growth. Our results lay the foundation for unveiling the roles of tRNA modifications in Mtb pathogenesis and developing new therapeutics against tuberculosis.

Introduction

tRNA is an adaptor molecule that enables protein synthesis by converting the triplet genetic code in mRNA into amino acids. The fidelity of base paring between mRNA codons and tRNA anticodons is monitored within ribosomes and is critical for properly incorporating the amino acids bound to the 3’ ends of tRNAs into growing polypeptides. For optimal translation, the abundances, and properties of tRNA isoacceptors are fine-tuned by diverse mechanisms, including chemical modifications[1–4] Dysregulation of tRNA abundance and/or structure leads to defective decoding and results in ribosome pausing and collisions, protein misfolding, stress responses and can have detrimental or lethal effects on the cell[5–10].

Chemical modifications of tRNA (tRNA modifications) are found in all kingdoms of life and fine-tune tRNA properties including mRNA decoding efficiency, recognition by aminoacyl-tRNA synthetases, half-life and structural stability[2, 10–12]. Modifications are prevalent in the anticodon loop, particularly at the first letter of the anticodon. Modifications of the anticodon loop directly modulate codon recognition, whereas modifications in the tRNA body region primarily stabilize tRNA tertiary structure, protecting them from degradation in the cell. tRNA modifications are generated by dedicated site-specific enzymes referred to as tRNA modifying enzymes. tRNA modifications have been extensively characterized in a few model organisms[2, 13], but their profiles, regulation, and functions in non-model organisms, including bacterial pathogens, are understudied[13]

tRNA sequencing (tRNA-seq) allows for the rapid and systematic prediction of many tRNA modification sites[14, 15]. We recently developed a comparative tRNA-seq protocol to profile tRNA modifications in organisms with uncharted tRNA modification profiles; in Vibrio cholerae, this approach led to the discovery of a new RNA modification and RNA editing process[16]. tRNA-seq enables rapid prediction of modified sites through detection of reverse transcription-derived signatures, such as nucleotide misincorporation and early termination, both of which occur more frequently at modified sites. Furthermore, several chemical treatments of tRNA can convert modifications that are not recognizable as reverse transcription-derived signatures into detectable signals, expanding the repertoire of modifications that can be distinguished by tRNA-seq[17–20].

Mycobacterium tuberculosis (Mtb), the agent of tuberculosis (TB), is a global pathogen that caused >10.5 million cases and over 1.5 million deaths worldwide in 2020[21]. Several studies have uncovered roles for non-Mtb mycobacterial tRNA modifications in stress responses, adaptation to environmental changes, and persister cell formation[22]. Mycobacterium bovis BCG, an organism closely related to Mtb, responds to hypoxia by reprogramming 40 ribonucleoside modifications in tRNA to facilitate translation of a subset of proteins that promote survival in hypoxic conditions[22]. To date, studies of the profiles and functions of Mtb tRNA modifications have been limited.

Here, we conducted tRNA-seq in Mtb. We assigned modifications to many of the reverse transcription-derived signatures identified, using information on the presence of homologs of known modifying enzymes in Mtb. Chemical treatments of tRNAs carried out prior to tRNA-seq increased the detectability of certain modifications. We constructed two Mtb deletion mutant strains, with deletions of mnmA and truB, and confirmed that the absence of these modification enzymes eliminated the predicted signals in tRNA-seq data. Furthermore, while deletion of mnmA in Mtb did not affect the pathogen’s growth in in vitro laboratory growth conditions, the mnmA knockout strain’s growth was attenuated in a macrophage infection model. Our findings suggest that tRNA modifications warrant further study as we unravel the complexity of Mtb infections, as they may serve as targets for new therapeutics.

Results

In silico prediction of Mtb tRNA modifying enzymes

To predict tRNA modifications in Mtb, we used Basic Local Alignment Search Tool (BLAST)[23] to identify homologues of all tRNA modification enzymes registered in Modomics in the Mtb genome[24] (Supplementary Table 1). With a stringent threshold (E-value <1×10−10), 31 Mtb genes homologous to genes encoding known RNA modification enzymes were identified (Supplementary Table 2). 23 of these genes are predicted to synthesize 16 tRNA modifications in Mtb (Supplementary Table 2 and 3), including miaA and miaB for 2-methylthio-6-isopentenyl-adenosine (ms2i6A), tsaD, tsaB, tsaE, and sua5 for N6-threonylcarbamoyladenosine (t6A), mnmA for 2-thiouridine (s2U), truB, truA, rluA, and pus9 for pseudouridine (Ψ), trm2 for 5-methyluridine (m5U), trmD for 1-methylguanosine (m1G), trmI for 1-methyladenosine (m1A), trmL for 2’-O-methylcytidine (Cm) or 2’-O-methyluridine (Um), trmB for 7-methylguanosine (m7G), two trmH for 2’-O-methylguanosine (Gm), trmR for 5-methoxyuridine (mo5U), trcM for 5-methylcytidine (m5C), dusB for dihydrouridine (D), tadA for inosine (I), and tilS for lysidine (k2C). While eight additional genes met the threshold for homology to modification enzymes, they exhibited greater similarity to a Glutamyl-tRNA synthase; Rv2992c, ribosome associated GTPases (Der; Rv1713 and Era; Rv2364c), molybdopterin biosynthesis proteins (MoeW; Rv2338c and MoeB2; Rv3116), thiosulfate sulfur transferases (CysA; Rv3117 and SseA; Rv3283) and a riboflavin biosynthesis protein (RibG; Rv1409), and likely do not correspond to tRNA modification enzymes (Supplementary Table 2).

We mined data from genome-wide Tn-seq and CRISPRi screens[25, 26] to assess the impacts of Mtb tRNA modifications on its growth. Five genes encoding Mtb tRNA modifying enzymes, trmD, tilS, tadA, tsaC2, and miaA, were reported to be essential for Mtb growth in both Tn-seq[25] and CRISPRi screens[26] (Supplementary Table 3), suggesting that the modifications they produce are critical for tRNA functions. Indeed, E. coli trmD, tilS, tadA, and tsaC2 are also essential for growth and critical for codon decoding, aminoacylation, and reading frame maintenance[27–30]. The Mtb modifications synthesized by these enzymes likely have similar impacts on tRNA functions. Unexpectedly, one modifying enzyme, MiaA, is non-essential in E. coli grown in nutrient rich medium, but apparently essential in Mtb, suggesting that i6A, the modification introduced by MiaA, may have more profound roles in Mtb translation than in E. coli.

Several of the putative Mtb tRNA modifying enzymes are conserved across all three domains of life (e.g., TruB and TsaD) (Fig. 1). By contrast, some enzymes were limited to bacterial species closely related to Mtb, possibly suggesting their species-specific physiological roles, including pathogenesis. For example, TrmI (Rv2118c) homologs are widely present in Archaea and Eukaryotes, but are sparsely distributed in bacteria, where they are primarily limited to Actinomycetia, including Mtb, and several thermophilic bacterial species (e.g., Thermus thermophilus). TrmI synthesizes m1A at position 58 in eukaryotes[31] and at position 57/58 in archaea[32]; indeed, TrmI in Mtb has been proven to generate m1A at position 58[33]. Pus9 (Rv3300c) has many homologs exhibiting weak similarity across eukaryotes and bacteria. However, some species, including Pseudomonas aeruginosa, Acinetobacter baumannii, Neisseria gonorrhoeae, and several species of Actinomycetia, encode proteins showing strong homology to Mtb Pus9, suggesting that this set of pseudouridylases have a distinctive property such as substrate specificity.

Phylogenetic distribution of Mtb tRNA modifying enzyme homologs.

Heat map of log10 E-values from BLAST search results. BLAST searches were conducted against 120 manually picked organisms using Mtb tRNA modifying enzymes as queries. When one organism has multiple hits, the lowest log10(Eval) values among hits is shown. iTol[34] was used to depict the results.

tRNA modifying enzymes predicted in Mtb were observed in mycobacteria species, including Mycolicibacterium smegmatis and Mycobacteroides abscessus (Fig. 1), suggesting that tRNA modification patterns are similar among mycobacterium species.

Profiling Mtb tRNA modification sites by tRNA sequencing

To begin profiling detectable Mtb tRNA modifications, we sequenced tRNAs isolated from wild type Mtb strain H37Rv grown in 7H9 medium. In this protocol, tRNAs are first reversed transcribed to cDNA[16]. During cDNA synthesis, chemical modifications on tRNA nucleotides disrupt Watson-Crick base pairing and increase the frequency of reverse transcriptase errors, leading to incorporation of the incorrect nucleotide or early termination of cDNA synthesis[35]. These reverse transcription derived ‘signatures’ typically correspond to modified sites[14, 15] and are depicted in the heat map in Fig. 2.

Heat map of misincorporation and early termination frequency in sequencing of tRNAs from wild type Mtb.

(A, B) Heatmaps show misincorporation (A) and termination (B) frequencies at all positions across tRNAs (read 5’ to 3’). Predicted modifications are labeled based on similarity to known modifications in other organisms and the presence of the tRNA modifying enzyme homologs (Supplementary Table 3). The positions with more than 10% misincorporation in Mtb but not in E. coli are depicted in white in A. (C) M. tuberculosis tRNA modifications predicted in this study. Schematic tRNA secondary structure with sites of modifications identified either by the presence of modifying enzymes and/or tRNA-seq. Modifications and tRNA species that are not observed in E. coli are shown in red. Modifications and positions that are predicted by both RT-derived signature and the presence of the homologs of tRNA modifying enzymes are shown in yellow (without chemical treatment) and green (with chemical treatment), whereas modifications that are only predicted by the presence of the homologs are shown in light blue. Genes reported to be essential in Mtb are shown in bold.

Comparison of the reverse transcription derived signatures observed in Mtb to E. coli, where tRNA modifications have been well characterized[16] enables the prediction of the presence of common modifications, including ms2i6A, m1G, I, and k2C (Fig 2 and Supplementary Fig. 1). These predictions are strongly supported by the set of tRNA modification enzymes identified in the Mtb genome (Fig. 1 and Supplementary Table 3), including miaA and miaB (ms2i6A), trmD (m1G), tadA (I), and tilS (k2C). Some tRNA modifications were observed in Mtb but are not present in E. coli. In other actinobacteria, A58 and A59 are likely modified to m1A[36], and since trmI, the methylase that generates this modification is present in Mtb[33], most Mtb tRNAs likely contain this modification as well. Nucleoside variation in the sequence of tRNA genes can account for some of the variations in modified sites between E. coli and Mtb. For example, in Mtb, termination signatures derived from G at position 37 were detected in tRNA-Arg2, -Gln1, and -Gln2, whereas this position in these tRNAs in E. coli are modified A, such as m2A and m6A, which are silent in tRNA-seq. Since m1G induces strong termination during reverse transcription, these positions are likely modified to m1G, as observed in the Bacillus subtilis tRNA-Arg2 gene position 37 G[37].

tRNA samples were also treated with several chemical treatments prior to sequencing, to expand the set of tRNA modifications detectable by tRNAseq. These treatments included iodoacetamide (IAA), for detection of sulfur modifications,1-cyclohexyl-(2-morpholinoethyl) carbodiimide (CMC) for detection of Ψ, and alkali for detection of D and m7G. The chemical treatment protocols were first carried out with E. coli, to validate the methods.

IAA is a thiol-reactive compound that covalently attaches carboxyamidomethyl to thiolated uridines via nucleophilic substitution[38] and modified s4U is detected as C instead of U. In IAA-treated samples, position 8 and 9, corresponding to s4U in many tRNA species, had high misincorporation frequencies, confirming that IAA treatment modifies s4U, leading to elevated misincorporation (Supplementary Fig. 2). Furthermore, we observed higher misincorporation and termination signals at the positions corresponding to other sulfur modification, s2C, s2U and their derivatives, such as position 32 in tRNA-Arg3, -Arg5, -Ser3, and Arg4, and 34 in tRNA-Glu, -Gln1, and -Lys (Fig. 3 and Supplementary Fig. 2), revealing that IAA treatment facilitates the detection of not only s4U but also additional sulfur modifications, which are weakly detected without the IAA treatment.

IAA treatment promotes detection of sulfur modifications on tRNAs by enhancing termination signals.

(A, B) Heatmaps of the termination signals of E. coli tRNAs treated with (A) or without (B) IAA. Known modification sites, including sulfur modifications (s4U, s2C, s2U in white) are shown. (C) Termination frequency at s4U, s2C, and s2U sites of tRNAs treated with or without IAA.

Next, we applied IAA treatment to Mtb tRNA-seq. IAA treatment increased termination signals from position 34 in tRNA-Glu1, -Gln1, and -Lys, which contain s2U derivatives in E. coli (Fig. 4A). The 2-thiouridine modification is carried out by MnmA in E. coli[39], and a homolog, Rv3024c, of this enzyme was identified in the Mtb genome (Supplementary Table 3)[40]. We used double stranded DNA-based recombineering to delete Rv3024c in Mtb, yielding strain MtbΔmnmA[41]. Sequencing of tRNA isolated from MtbΔmnmA with prior IAA treatment showed reduced termination signals from position 34 in tRNA-Glu1, -Gln1, and -Lys in treated samples, indicating that Rv3024c plays a critical role in the modification responsible for increased termination frequency derived from position 34 in these tRNAs (Fig. 4B and 4C). Together, these observations strongly suggest that Rv3024c encodes an MnmA-like enzyme that sulfurates position 34 uridines in three Mtb tRNA isoacceptors.

Heat map and plot of early termination frequency from sequencing of tRNAs from wild type and MtbΔmnmA with and without RNA alkylation.

(A, B) Heat map of early termination frequencies across tRNA molecules and positions for WT (A) and MtbΔmnmA (B). Sulfur modification is shown in white. (C) Plot of termination frequencies at position 37 in wild type (WT) Mtb and MtbΔmnmA for lysine_UUU, glutamate_UUG, and glutamine_UUC isoacceptors. IAA; iodoacetamide.

As shown previously[42], CMC treatment increased both misincorporation and termination signatures at a subset of Ψs in E. coli tRNAs (Fig 5, and Supplementary Fig. 3 and 4). In addition, CMC-treated samples showed increased frequencies of both misincorporation and termination at position 16, 17, 20, and 20A corresponding to D, and m7G at position 46. Both modifications are known to undergo base elimination in mild alkali conditions[43]. Since these signals were also observed in the reaction condition in which CMC was not added, these signals are likely attributable to the alkali treatment that is common to both conditions.

CMC and alkali treatment facilitate detection of Ψ, D, and m7G modifications in E. coli.

(A, B) Heatmaps of the misincorporation signals of E. coli tRNAs treated with (A) or without (B) CMC. In both conditions, tRNAs are incubated in alkali condition. Known Ψ, D, and m7G sites are shown. Ψ is shown in white. (C) Misincorporation frequency at known Ψ, D, m7G sites of tRNAs treated with CMC+/alkali+ or CMC-/alkali+, and tRNAs without treatment.

Reverse transcription-derived signatures derived from alkali-treated D showed a distinctive pattern. With this treatment, when two Ds are at consecutive positions, e.g., D16 and D17, termination signals were elevated at the following position, i.e., position 18 (Supplementary Fig. 3, 5 and 6). Furthermore, alkali treatment also led to higher misincorporation frequencies at singlet Ds (Fig. 5, Supplementary Fig. 5 and 6). Thus, termination and misincorporation signatures enabled the prediction of known E. coli tRNA sites modified to D.

CMC/alkali treatment facilitated the identification of additional modifications in Mtb tRNAs. Regardless of CMC treatment, alkali treated samples showed increased misincorporation and termination frequencies derived from U located at position 16, 17, 20, and 20A (Fig. 6 and Supplementary Fig. 7). As observed in E. coli, termination signals at position 18 and 21 likely correspond to consecutive Ds at position 16 and 17, and 20 and 20A, respectively. Mtb Rv0823c is a homolog of dihydrouridylase DusB (Supplementary Table 3), which likely accounts for the synthesis of D at these positions. Furthermore, alkali treatment increased the misincorporation frequencies at G at position 46 (Supplementary Fig. 8). Since Rv0208c is a homolog of TrmB, which synthesizes m7G at position 46 in E. coli, multiple Mtb tRNA species likely contain m7G at position 46 (Fig. 6 and Supplementary Fig. 8).

Heat map of early termination frequency from sequencing of tRNAs isolated from WT and MtbΔtruB following CMC treatment.

Heat map of early termination frequencies across tRNA molecules and positions for WT (left) and MtbΔtruB (right). Termination signals derived from position 55 are shown in white.

CMC treatment also increased the termination frequency at several sites. Termination signatures derived from position 55, which is exclusively uridine in all tRNA species, increased in most tRNA species, suggesting that Mtb tRNAs contain pseudouridines at this position[37]. Rv2793c is an Mtb homologue of E. coli TruB and deletion of Rv2793c reduced the termination frequencies at this position in tRNAs isolated from MtbΔtruB (Fig. 6 and Supplementary Fig. 9)[41]. Together, these observations suggest that Rv2793c encodes a TruB-like enzyme that modifies position 55 uridines to Ψ across tRNA species. Furthermore, the presence of a TruA homolog in Mtb (Rv3455c) suggests that in multiple tRNA species U at positions 38-40 can be modified to Ψ. Indeed, the termination signatures derived from positions 38 and 39 increased depending on CMC treatment, strongly suggesting that these positions are modified to Ψ (Fig. 6 and Supplementary Fig. 9).

In total, among 16 tRNA Mtb tRNA modifications predicted by the presence of tRNA modifying enzymes (Fig 1 and Supplementary Table 3), 9 species of modifications were detected based on reverse-transcription derived signatures (Fig. 2C).

Growth of MtbΔmnmA is attenuated in a macrophage infection model

To address whether Mtb tRNA modifications impact the pathogen’s growth in the host environment, we used the MtbTnDB transposon insertion sequencing (Tn-seq) database[44, 45] to determine if transposon insertions in genes encoding tRNA modifying enzymes have been associated with in vivo growth defects. Transposon insertions in mnmA were reported to attenuate Mtb growth in mice infected with a library of Mtb transposon mutants, suggesting that mnmA facilitates Mtb growth in vivo. We found that the growth of WT and MtbΔmnmA were similar in 7H9 medium (Fig. 7), suggesting the absence of s2U modification at position 34 does not impair Mtb growth in culture. In contrast, the MtbΔmnmA mutant was significantly impaired for growth in a macrophage infection model[45] (Fig. 7). Defective growth of the MtbΔmnmA mutant was also observed in macrophages treated with all-trans retinoic acid (ATRA), which promotes macrophage control of Mtb infection[46]. These observations strongly suggest that modification of U to s2U by MnmA facilitates Mtb growth in macrophages.

MtbΔmnmA is attenuated in a macrophage infection model.

Top: Wild type and MtbΔmnmA do not display growth differences in 7H9 medium. Bottom: Auto-luminescent wild type and ΔmnmA Mtb strains were diluted to a multiplicity of infection of 2 bacteria per mouse bone marrow-derived macrophage with or without all-trans retinoic acid (ATRA). Survival was measured by luminescence and normalized to luminescence at time 0.

Discussion

Here, we profiled Mtb tRNA modifications with tRNA sequencing to provide the first maps of the tRNA modification landscape in this global pathogen. In total, nine modifications, including six modifications without chemical treatment, were identified based on reverse transcription derived signatures. CMC/alkali treatment and IAA treatment further identified Ψ and m7G and sulfur modifications, respectively. Although we did not chemically validate the modifications predicted by tRNA-seq with mass spectrometry, the identification of Mtb homologs of tRNA modifying enzymes strongly bolsters the RT-signature-based predictions. Furthermore, the deletion of truB and mnmA genes in Mtb eliminated the respective modification signatures of pseudouridine and s2U, validating that the enzymes encoded by these genes synthesize these modifications. Finally, the growth defect of the ΔmnmA strain within macrophages but not in a nutrient-rich medium suggests that s2U tRNA modification facilitates Mtb adaptation to the host intracellular environment.

IAA treatment was developed for detecting s4U modification in pulse-chase experiments to measure RNA turnover[38]. We found that this treatment enhances the RT-signatures of additional sulfur modifications as well, including s2C and s2U, indicating that IAA should have general utility in tRNA-seq-based profiling of sulfur modifications, including in studies tracking changes in tRNA sulfuration[47, 48].

Multiple modifications do not cause reverse transcriptase errors. Although tRNA-seq provides a simple and rapid method for profiling the landscape of modifications in all tRNA species, this approach does not enable comprehensive identification of modifications. Here, several modifications, including t6A, m5U, mo5U, m5C, Cm, Um, and Gm, predicted by the presence of Mtb homolog of tRNA modifying enzymes, such as TsaC2, TsaB, TsaD, TsaE, Trm2, TrcM, TrmR, TrmL, and TrmH, were not detected in tRNA-seq. Additionally, CMC treatment may not determine all Ψ positions, thus, the target of Rv3300 and Rv1540 remain unclear. Since these genes are similar to E. coli rluA, which also targets rRNA, these genes may target rRNAs instead of tRNAs. Mapping and further analysis of these modifications will require different approaches such as tRNA purification and RNA mass spectrometric analysis.

Several strong RT signatures were detected in Mtb tRNAs but not in E. coli (Fig. 2A), including G45 in tRNA-Gly2. Most of these signals are strict misincorporations of C or T at A or G position, respectively. Since modifications at a purine position, e.g., m1G and m1A, generally cause random misincorporation of the other three nucleosides, these Mtb-specific misincorporation signatures are not likely to be derived from modification. Further mass spectrometric analysis of purified tRNAs will be necessary to assess whether these signals are indeed derived from Mtb-specific modifications; such Mtb-specific modifications could be related to Mtb pathogenesis and/or survival in the host.

The role of s2U in Mtb appears to be unusual. s2U is a universally conserved modification observed in all three domains of life at the wobble position in the anticodon. This modification enhances the stacking of s2U with U35 to stabilize the anticodon structure, facilitating codon-anticodon interactions. s2U is also recognized by multiple aminoacyl-tRNA synthetases for efficient amino acylation[12]. Although elimination of this sulfur modification causes severe growth phenotypes in most organisms[39, 49], unexpectedly, the deletion of mnmA in Mtb did not attenuate growth of the pathogen in vitro, suggesting that the Mtb requirement for s2U modification differs from other organisms. The marked growth retardation of ΔmnmA strain within macrophages indicates the specific requirement of this modification within host cells. This modification may be necessary for maintaining general translation efficiency inside host cells and/or facilitate the expression of specific genes that are necessary for survival within macrophages. In fact, lack of mnmA is reported to sensitizeE. coli to oxidative stress, raising the possibility that s2U modification promotes Mtb growth under oxidative stress elicited by the host.

Differential codon usage between house-keeping genes and virulence genes could contribute to the differential growth phenotypes observed in vitro and in vivo in the ΔmnmA mutant. Among multiple codons encoding Glu, Gln, and Lys, s2U modification dependent codon usage might be preferentially distributed in genes associated with intracellular growth. For example, Mtb has two tRNA isoacceptors, tRNA-Glu1(s2UUC) and tRNA-Glu2(CUC), to decipher two Glu codons, GAA and GAG. According to the wobble paring rule, GAA is only decoded by tRNA-Glu1(s2UUC), whereas GAG is decoded by both tRNA-Glu1(s2UUC) and tRNA-Glu2(CUC); i.e., GAG can be deciphered by an s2U-independent tRNA. Thus, genes required for intracellular growth might be enriched with GAA, an s2U-dependent codon. Similar codon usage differences could be present in Gln and Lys codons deciphered by s2U-containing tRNAs.

In most organisms, s2U is further modified into derivatives containing an additional chemical moiety at position 5[2, 50, 51]. However, Mtb does not contain apparent homologs of the tRNA modifying enzymes that introduce the additional modifications to s2U. Thus, Mtb may contain s2U or s2U derivatives synthesized by other types of enzymes. Additional analyses to elucidate the structures of modified s2U in Mtb are warranted.

Another unexpected finding is the presence of a TrmR homolog in Mtb. TrmR in Bacillus subtilis is a methylase that converts a 5-hydroxyuridine (ho5U) into a 5-methoxyuridine (mo5U) at the tRNA wobble position, suggesting that Mtb possesses mo5U. However, Mtb does not encode an apparent homolog for the enzyme that mediates the RNA hydroxylation reaction that yields ho5U, the substrate for TrmR. Although we cannot rule out the possibility that Mtb TrmR methylates a substrate other than RNA, it is possible that Mtb utilizes an unknown hydroxylation pathway for synthesizing ho5U.

Our findings serve as a valuable starting point for the research community to continue characterizing the physiological roles and mechanisms of Mtb tRNA modifications. Since tRNA-seq offers an efficient and scalable platform for surveying changes in tRNA modifications, this approach will be valuable across growth conditions, and may be extended to growth inside host cells. Finally, further studies elucidating the mechanisms by which tRNA modifications facilitate Mtb growth in host cells should be valuable for designing new therapeutics for tuberculosis.

Supporting information